Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-06-16 analysis/memo2_TAS_reviewers.Rmd
response to extreme incompatibility scenario.

By "cognizable" class, we mean the social category to which the struck juror belongs that is the possible basis of illegal strike bias. For example, if the set of prospective jurors consists only of Black and White individuals and the focus is on strike bias _against_ Black prospective jurors because of their race, the "cognizable" class is "Black prospective jurors". If, however, the focus is on strike bias against Black female prospective jurors because of their race and gender, the "cognizable" class is "Black female prospective jurors." We used the phrase "cognizable class members subject to strike" instead of "cognizable class jurors", because at the time any strike occurs, (1) no one is actually a juror, because no one has yet been seated onto the jury, and (2) not every prospective juror could have been struck.

@@

Subsetting the historical data does not necessarily reduce the degree of incompatibility between the historical strike data and the current trial ($b_hist$ - $b_curr$). As we discuss in the paper (page X, lines YY-YY), it depends on what one assumes about how much the subsetting criteria (e.g., only "similar" cases) changes the degree of incompatibility. We stress "assumes", because the true values of $b_hist$ and $b_curr$ are not directly observable in actual cases. Accordingly, for any subset of the historical strike data one chooses, our guidance for choosing an $\alpha$ value remains the same: Check for how sensitive to the posterior estimate of strike bias to different values of $\alpha$. The current version of the software prototype does not let users subset the historical data by some characteristic (e.g., date, defendant race). However, in principle, it could be modified to add that kind of feature.

@@

First, we agree that high incompatbility may lead to error in bias detection. The challenge is that high incompatibility (in the case where $b_curr = 0$ or where $b_hist = 0$) is not directly observable.  This is why we recommend using sensitivity to choice of $\alpha$ value to ascertain whether to worry about high incompatability.

Second, we wish to address the idea that a judge may be reluctant to "seriously entertain a Batson challenge" if the strikes in the present case indicate essential equal "removal rates of protected and unprotected members of the venire." we take "seriously entertain" to denote the premise that judge would be less likely to find a _Batson_ prima facie case and ultimately find a _Batson_ violation, given those observed "removal rates". We do not dispute that premise about judge behavior.

>>>>

By "cognizable" class, we mean the social category to which the struck juror belongs that is the possible basis of illegal strike bias. For example, if the focus is on strike bias against Black prospective jurors because of their race, the "cognizable" class is "Black prospective jurors". If, however, the focus is on strike bias against Black female prospective jurors because of their race and gender, the "cognizable" class is "Black female prospective jurors."

We use the phrase "cognizable class members subject to strike" instead of "cognizable class jurors", for two reasons. First, at the time any strike occurs, no one is a "juror", because no one has yet been seated onto the jury. Second, at the time any strike occurs, not every prospective juror could have been struck. As we assume in the paper (see pages XXX), only a subset of prospective jurors present during jury selection are ever subject to strike at any one time. The size of that subset is usually based on the number of seats of the jury itself plus the total number of strikes that the parties could exercise.

@@

Subsetting the historical data does not necessarily reduce the degree of incompatibility between the historical strike data and the current trial ($b_hist$ - $b_curr$). As we discuss in the paper (page X, lines YY-YY), it depends on what one assumes about how much the subsetting criteria (e.g., the salient criteria for "similar" cases) changes the degree of incompatibility. We stress "assumes", because the true values of $b_hist$ and $b_curr$ are not directly observable in actual cases. Accordingly, for any subset of the historical strike data one chooses, our guidance for choosing an $\alpha$ value remains the same: Check for how sensitive the posterior estimate of illegal strike bias is to different values of $\alpha$. The current version of the software prototype does not let users subset the historical data by some characteristic (e.g., date, defendant race). However, in principle, it could be modified to add that kind of feature.

@@

We take this comment to posit that a judge is unlikely to find ("seriously entertain") even a _Batson_ prima facie case if the posterior estimate of $b$ with $\alpha = 0$ detects no illegal bias against the cognizable class (e.g. Black prospective jurors), because $b = 0$ falls within the credible interval, even if the posterior estimate of $b$ given some set of non-zero values of $\alpha$ detects illegal bias, i.e., $b = 0$ falls outside the credible interval.  This might occur, for example, where an attorney is motivated to strike Black prospective jurors because they are Black _only_ in cases with Black defendants; all that attorney's past trials involved Black defendants; but the current trial has a White defendant. In this extreme scenario, we agree that a judge might properly worry about high incompatibility and not find a _Batson_ prima facie case for that reason, even though no one can directly observe or know with complete certainty whether this is a case where $b_curr = 0$ and $b_hist >> 0$. Or the judge may still find enough for a prima face case, but downweight the historical data when ultimately deciding whether illegal bias likely occurred. In the reverse extreme situation, the posterior estimate of $b$ with $\alpha = 0$ detects illegal bias but does not detect illegal bias when $\alpha > 0$. Here, too, we agree that a judge might properly worry about high incompatibility, except that down-weighting the historical data here increases the support for finding at least enough a _Batson_ prima facie case.
